|

### 2023-04-04 paper.Rmd
fixing narrative in Background section 3

One may argue that reading diagnostic plots is to some extent subjective and indecisive compared to those rigorous statistical procedures as it relies on graphical perception - human ability to interpret and decode the information embedded in graph [@cleveland_graphical_1984]. Further, the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. For instance, people 
over-interpret the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015].

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots. Later, @buja_statistical_2009 proposed the lineup protocol as a visual test inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ null plots have the identical graphical procedure except the data has been replaced with data consistent with $H_0$, that the model is correctly specified. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$.

>>>>

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in graph [@cleveland_graphical_1984]. Such examination thus feels there is a subjective nature to evaluate, say, diagnostic plots to infer appropriateness of statistical models, although arguably we would expect such evaluation should be done objectively and such human-to-human variation in "reading" data plots should be minimised.

In practice, over-interpretation of a single plot is common.  For instance, in @roy_chowdhury_using_2015, some over-interpreted the separation between gene groups in a two-dimensional projection of a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups. To mitigate this over-interpretation, @buja_statistical_2009 proposed a line-up protocol to assess plots in a manner analogous to the null hypothesis significance testing framework. More specifically, the protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ null plots have the identical graphical procedure except the data has been replaced with data consistent with $H_0$, that the model is correctly specified. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$.
